---
title: 'Probability Practice Problems'
date: '2021-09-10'
slug: probability
categories: []
tags: []
header-includes:
  - \usepackage{xfrac}
bibliography: 'citations.bib'
csl: journal-of-mathematical-physics.csl
# mathematical-medicine-and-biology.csl
nocite: '@*'
---

<script src="https://cdnjs.cloudflare.com/ajax/libs/prism/0.0.1/prism.min.js"></script>
<script src="//cdnjs.cloudflare.com/ajax/libs/highlight.js/10.5.0/highlight.min.js"></script>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
library(kableExtra)
library(knitr)
library(ggplot2)
```



<div class="definition1">
Exercises from the book Introduction to Probability by Joseph K. Blitzstein and Jessica Hwang &copy; Chapman & Hall/CRC Press, 2014 [@blitzstein-2019].
</div>


## Sample Spaces 

### Problem 0.1 

Sara and Megan are two students who show up for a standardized test. Sara has a $\tfrac{1}{8}$ chance of passing the test, whereas Megan has a $\tfrac{5}{12}$ chance. What are the chances that at least one of them will pass the test?

<div class="example1">
**Solution.** Let $A$ denote the event that Sara passes the test and $B$ denote the event that Megan passes. Then we have that $P(A) = \tfrac{1}{8}$ and $P(B)=\tfrac{5}{12}$. Now, the probability of at least one of them passing the test is denoted as the union of $A$ and $B$, which means

$$
\begin{align}
P(A \cup B) & \equiv P(A) + P(B) - P(A \cap B) \\
& \equiv P(A) + P(B) - P(A)P(B) \\
&\equiv \tfrac{1}{8} + \tfrac{5}{12} - \tfrac{1}{8}\times \tfrac{5}{12} \\
&\equiv \tfrac{47}{96}
\end{align}
$$


</div>



------------------------------------------




## Conditional Probability and Bayes' Rule



### Problem 1.1 


An email filters spam by scanning commonly occurring words in spam. Suppose that $70\%$ of email is spam. In $10\%$ of the spam emails, the expression "fast cash" appears, whereas this phrase only appears in $1\%$ of non-spam emails. A new email has just arrived, which mentions "fast cash" multiple times. What is the probability that it is indeed spam?


<div class="example1">

**Solution.** Let $S$ be the event that the mail is spam and $E$ be the event that the mail contains the phrase "fast cash". We want to find $P(S|E)$. We have that $P(E|S) = 0.10$, $P(E|S^c) = 0.01$, $P(S) = 0.70$, and $P(S^c) = 0.30$. By the law of total probability, 


$$
\begin{align}
P(E) & \equiv P(E|S)P(S) + P(E|S^c)P(S^c) \\
& \equiv 0.10 \times 0.70 + 0.01 \times 0.30 \\
& \equiv 0.073
\end{align}
$$

Applying Bayes' rule, we have

$$
\begin{align}
P(S|E) & \equiv \dfrac{P(E|S)P(S)}{P(E)} \\
&\equiv \dfrac{0.10 \times 0.70}{0.073} \\
& \equiv 0.9589
\end{align}
$$



</div>



-----------------------------------------------


### Problem 1.2

Companies $A, B, C$ manufacture screens used for a particular type of computer. The supplied proportions of screens for companies $A, B, C$ are $0.5, 0.3, 0.2$, such that $0.01, 0.02, 0.03$ are defective, respectively. Given that the screen on such a computer is defective, what is the probability that Company $A$ manufactured it?


<div class="example1">

**Solution.** Let $C_1$ be the event that company $A$ manufactured the screen and let $D$ be the event that the screen is defective. We want to find $P(C_1|D)$. We're given that $P(C_1)= 0.5$, $P(C_2)= 0.3$, $P(C_3)= 0.2$, $P(D|C_1) = 0.01$, $P(D|C_2) = 0.02$, and $P(D|C_3) = 0.03$. Using Bayes' rule and the law of total probability:


$$
\begin{align}
P(C_1|D) & \equiv \dfrac{P(D|C_1)P(C_1)}{P(D)} \\
& \equiv \dfrac{P(D|C_1)P(C_1)}{P(D|C_1) P(C_1) + P(D|C_2) P(C_2) + P(D|C_3) P(C_3)} \\
& \equiv \dfrac{0.01 \times 0.5}{0.01 \times 0.5 + 0.02\times 0.3 + 0.03 \times 0.2} \\
& \equiv 0.2941176
\end{align}
$$

</div>



--------------------------------


### Problem 1.3 

A family has $3$ children, denoted $A, B, C$. Find the probability that $A$ is older than $B$, given that $A$ is older than $C$. 

<div class="example1">

**Solution.** We want to find $P(A \gt B \,{\mathbf |}\, A \gt C)$. The possible birth orders are $ABC$, $ACB$, $BAC$, $BCA$, $CAB$, and $CBA$. Therefore, we have

$$
\begin{align}
P(A \gt B | A \gt C) &\equiv 
\dfrac{P \left(A\gt B  \bigcap  A\gt C \right)}{P(A \gt C)} 
\equiv \dfrac{{1}/{3}}{{1}/{2}} 
\equiv \frac{2}{3}
\end{align}
$$

</div>



--------------------------------


### Problem 1.4 


Consider an election with candidates $A$ and $B$. For a random voter, let $A$ be the event they voted for $A$, and $W$ be the event they're willing to participate in an exit poll. Suppose that $P(W|A) = 0.7$ but $P(W|A^c) = 0.3$. In the exit poll, $60\%$ of the respondents report they voted for $A$. Find $P(A)$, the true percentage of people who voted for $A$. 

<div class="example1">

**Solution.** To find $P(A)$, we use Bayes' rule to obtain $P(A|W)$:

$$
\begin{align}
P(A | W) & \equiv \dfrac{P(W|A)P(A)}{P(W)}
\end{align}
$$


Let $P(A) = p$. We're given $P(W|A) = 0.7$ and $P(A|W) = 0.6$. To obtain $P(W)$, the probability the voter participates in the exit poll, we use the law of total probability:

$$
\begin{align}
P(W) &\equiv P(W|A)P(A) + P(W|A^c)P(A^c) \\
&\equiv 0.7 p + 0.3\times (1-p) \\
&\equiv 0.4p + 0.3
\end{align}
$$


Plugging in $P(W) \equiv 0.4p + 0.3$ into $P(A|W)$,

$$
\begin{align}
P(A | W) & \equiv \dfrac{P(W|A)P(A)}{P(W)} \\
0.6 & \equiv \dfrac{0.7p}{0.4p + 0.3} \\
0 &\equiv 0.46p - 0.18 \\
p &\equiv 0.39
\end{align}
$$



Hence, the value of $P(A)$ is $0.39$.

</div>


-----------------------------------

### Problem 1.5


A student answers a multiple-choice problem and must choose one of the $n$ choices. Let $K$ be the event in which they know the answer, and $R$ be the event where they choose the correct answer. Suppose that if they know the correct answer, they'll get the problem correct. But if they don't know, they'll guess. Let $P(K) = p$. Find $P(K|R)$ in terms of $p$ and $n$. 

<div class="example1">

**Solution.** We're given $P(R|K) =1$. Whereas, if they don't know the correct answer, they must randomly guess out of the $n$ choices, i.e., $P(R|K^c) = \tfrac{1}{n}$. Using Bayes' Rule and the law of total probability, 


$$
\begin{align}
P(K|R) &\equiv \dfrac{P(R|K)P(K)}{P(R)} \\
&\equiv \dfrac{p}{P(R|K)P(K) + P(R|K^c)P(K^c)} \\
&\equiv \dfrac{p}{p + \tfrac{1}{n}(1-p)} \\
\end{align}
$$
</div>


-----------------------------------

### Problem 1.6 


There are $100$ coins in a jar, of which $99$ are fair and $1$ is double-ended. Suppose we toss a coin $7$ times, and all $7$ times land on $H$. Find the probability that the coin is double-headed.

<div class="example1">

**Solution.** Let $E$ be the event that the coin lands on $H$ all $7$ times and $D$ be the event that the chosen coin is double-headed. We want to find $P(D|E)$. 

$$
\begin{align}
P(D|E) & \equiv \dfrac{P(E|D)P(D)}{P(E)} \\
& \equiv \dfrac{P(E|D)P(D)}{P(E|D)P(D) + P(E|D^c)P(D^c)} \\
& \equiv \dfrac{\tfrac{1}{100}}{\tfrac{1}{100} + \left(\tfrac{1}{2}\right)^7 \cdot \tfrac{99}{100}} \\
&\equiv 0.5638765
\end{align}
$$

</div>


-----------------------------------


## Discrete Random Variables



### Problem 2.1 



Suppose we toss a fair coin twice. The sample space consists of four possible results: $\{HH, HT, TH, TT\}$. Let the random variable $X$ be the number of heads. Find the PMF of $X$.

<div class="example1">

**Solution.** For the random variable $X$, the possible values are $0, 1, 2$:

$$
\begin{gather}
X(HH) = 2, & X(HT) = X(TH) = 1, & X(TT) = 0
\end{gather}
$$

The PMF of $X$ is the function $p_X$ given by


$$
\begin{align}
p_X(0) & \equiv P(X=0) \equiv \tfrac{1}{4}, \\
p_X(1) & \equiv P(X=1) \equiv \tfrac{1}{2}, \\
p_X(2) & \equiv P(X=2) \equiv \tfrac{1}{4},
\end{align}
$$

and $p_X(x) \equiv 0$ for all other values of $x$. 


```{r, out.width=300, fig.align='center'}
PMF <- data.frame(x=c(0,1,2),y=c(0.25,0.5,0.25))
ggplot(PMF, aes(x=x, y=y)) +
  ylim(0, 0.75) + xlim(-1,3) +
  geom_segment( aes(x=x, xend=x, y=0, yend=y), color="black") +
  geom_point( color="#3291FC", size=4) +
  theme_light() + theme(panel.grid = element_blank()) +
  xlab("X") + ylab("PMF")
```


</div>



----------------------------


### Problem 2.2

In a tennis match, the contestants play $10$ games. In each game, the probability of a player winning is $0.4$, and the probability of a tie is $0.6$. Find the probability of exactly $5$ games ending in a tie. 

<div class="example1">
**Solution.** Let $T$ be the number of games ending in a tie. Then $T\sim \mathrm{Bin}(10,0.6)$, so the PMF is 

$$
\begin{align}
P(T=t) & \equiv 
\begin{pmatrix}10 \\ t\end{pmatrix} 
0.6^t \cdot 0.4^{n-t} \quad \text{for } t = 0,1,\ldots , n \\
P(T=5) & \equiv 
\begin{pmatrix}10 \\ 5\end{pmatrix} 
0.6^5 \cdot 0.4^{10-5} \\
& \equiv \frac{10!}{5! \cdot 5!}\cdot 0.6^5 \cdot 0.4^5 \cong 0.20
\end{align}
$$

</div>



---------------------------------


### Problem 2.3

You have a bag with $6$ marbles. One marble is red. You draw from the bag $100$ times, with replacement. What is the probability of drawing a red marble at least once?

<div class="example1">
**Solution.** Suppose $X$ is number of times a red marble is drawn. Then we have $X \sim \mathrm{Bin}\left(n=6, p=\tfrac{1}{6}\right)$. By the naive definition of probability, this is 

$$
\begin{align}
\mathbf{P}(X \geq 1) & \equiv 1 - \mathbf{P}(x=0) \\
&\equiv 1 - {\small\begin{pmatrix}100 \\ 0 \end{pmatrix}}
\left(\tfrac{1}{6}\right)^{0} \left(\tfrac{5}{6}\right)^{100} \\
& \equiv 1 - \left(\tfrac{5}{6}\right)^{100} \\
& \approx 0.99
\end{align}
$$



</div>


-------------------------------------------


## Continuous Random Variables


-------------------------------------------

### References
